Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-08-15 manuscript/draft.Rmd
more anna comments

**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
function in @forecast1] to represent a array of different time-series modeling
approaches. These models can include an autoregressive component (as in the
"naive" model, but with the possibility of longer-term dependencies in the
underlying process), a moving average component (where the noise can have serial
autocorrelation) and an integration/differencing component (so that the analysis
could be performed on sequential differences of the raw data, accommodating more
complex patterns including trends). The `auto.arima` function chooses whether to
include each of these components (and how many terms to include for each one)
using AICc [@forecast1]. Since there is no seasonal component to the BBS
time-series, we did not include a season component in these models. Otherwise we
used the default settings for this function.

### Models: continental-scale models

In contrast to the site-level time-series models, most attempts to predict
species richness focus on using correlative models based on environmental
variables. We tested three common variants of this approach: direct modeling of
species richness; stacking individual species distribution models; and joint
species distribution models (JSDMs). Following the standard approach, site-level
random effects were not included in these models as predictors, meaning that
this approach implicitly assumes that two sites with identical Bioclim,
elevation, and NDVI values should have identical richness distributions. As
above, we included observer effects & the associated uncertainty by running
these models 500 times (once per MCMC sample).

**"Macroecological" model.** We used a boosted regression tree model using the
`gbm` package [@ridgeway2017] to directly model species richness as a function
of environmental variables. Boosted regression trees are a form of tree based
modeling that work by fitting thousands of small tree-structured models
sequentially, with each tree optimized to reduce the error of its predecessors.
They are flexible models that are considered well suited for prediction
[@elith2008]. This model was optimized using a Gaussian likelihood, with a
maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000 trees. The
number of trees used for prediction was selected using the "out of bag"
estimator; this number averaged `r numbers$n_gbm[1]` for the non-observer data
and `r numbers$n_gbm[2]` for the observer-corrected data.

**Species Distribution Model.** Species distribution models (SDMs) predict
individual species' occurrence probabilities using environmental variables.
Species-level models are used to predict richness by summing the predicted
probability of occupancy across all species at a site. This avoids known
problems with the use of thresholds for determining whether or not a species
will be present at a site [@calabrese2014; pellissier2013]. Following
@calabrese2014, we calculated the uncertainty in our richness estimate by
treating richness as a sum over independent Bernoulli random variables:
$\sigma^2_{richness} = \sum_i{p_i (1 - p_i)}$, where $i$ indexes species. By
itself, this approach is known to underestimate the true community-level
uncertainty because it ignores the uncertainty in the species-level probabilites
[@calabrese2014]. To mitigate this problem, we used an ensemble of 500 estimates
for each of the species-level probabilities instead of just one, propagating the
uncertainty forward. We obtained these estimates using using random forests, a
common approach in the species distribution modeling literature based on fitting
hundreds of independent regression trees to randomly-perturbed versions of the
data [@cutler2007; @caruana2008]. When correcting for observer effects, each of
the 500 trees in our species-level random forests used a different Monte Carlo
estimate of the observer effects as a predictor variable.

**Joint Species Distribution Models.** Joint species distribution models (JSDMs)
are a new approach that makes predictions about the full composition of a
community instead of modeling each species independently as above. JSDMs remove
the assumed independence among species and explicitly account for the
possibility that a site will be much more (or less) suitable for birds in
general (or particular groups of birds) than one would expect based on the
available environmental measurements alone. As a result, JSDMs do a better job
of representing our uncertainty about richness, while stacked SDMs underestimate
it [@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
[@harris2015] because it is the only JSDM that describes species' environmental
associations with nonlinear functions.

@@

Variance in species richness within sites was lower than predicted by several
common count models, such as the Poisson or binomial (i.e. richness was
underdispersed for individual sites), so these count models would have had
difficulty fitting the data. Since richness had a relatively large mean 
(`r numbers$richness_summary$mean`) and all models already produce continuous
richness estimates (due to summing probabilities in the (J)SDM case), we defined
model performance for all models in terms of continuous Gaussian errors instead
of using discrete count distributions. When a model was run multiple times for
the purpose of correcting for observer effects, we used the mean of those runs'
point estimates as our final point estimate and we calculated the uncertainty
using the law of total variance (i.e.  the average of the model runs' variance,
plus the variance in the point estimates).

We evaluated each model's for forecasts using the data for each year between `r
timeframe$last_train_year` and `r timeframe$end_yr`. We used three metrics for
evaluating performance: 1) root-mean-square error (RMSE) to determine how far,
on average, the models' predictions were from the observed value; 2) the 95%
prediction interval coverage to determine how well the models predicted the
range of possible outcomes; and 3) deviance (i.e. negative 2 times the Gaussian
log-likelihood) as an integrative measure of fit incorporating good point
estimates, precision, and coverage.  In addition to looking at forecast
performance in general we evaluated how performance changed as the time horizon
of forecasting increased by plotting performance metrics against the forecast
horizon. Finally, we decomposed each model's squared error into two components:
one described the squared error associated with site-level means, while the
other described the squared error associated with annual fluctuations in
richness within a site. This decomposition tells us the extent to which each
model's error depends on consistent differences among sites versus changes in
site-level richness from year to year.

All analyses were conducted using R [@R]. Primary R packages used in the
analysis included dplyr [@dplyr], tidyr [@tidyr], gimms [@gimms], sp [@sp1;
@sp2], raster [@raster], prism [@prism], rdataretriever [@rdataretriever],
forecast [@forecast1; @forecast2], git2r [@git2r], ggplot [@ggplot2], mistnet
[@harris2015], viridis [@viridis], rstan [@rstan], yaml [@yaml], purrr [@purrr],
gbm [@gbm], randomForest [@randomForest]. Code to fully reproduce this analysis
is available on GitHub (https://github.com/weecology/bbs-forecasting) and
archived on Zenodo ().

@@

When comparing forecasts for richness across sites all methods performed well
(Figure 4; all $R^2 > 0.5$). However SDMs (both stacked and joint) and the
macroecological model all failed to successfully forecast the highest-richness
sites resulting in a notable clustering of predicted values near ~60 species and
the poorest performance ($R^2$=`r numbers$env_R2s`, versus $R^2$=`r numbers$ts_R2s` 
for the within-site methods).

![Performance of the six models we evaluated when predicting one year into the
future (2004) or ten years into the future (2013).  In general, the time series
models (**A.**) outperformed the environmental models (**B.**), and the accuracy
of the predictions declined as the forecast horizon was
extended.](../figures/scatter.png)

While the models generally did well in absolute terms (Figure 4), none
consistently outperformed the "average" baseline from the generalized linear
mixed model (Figure 4). The auto-ARIMA was generally the
best-performing non-baseline model, but in many cases (67% of the time), the
auto.arima procedure automatically selected a model with no autoregressive,
drift, or moving average terms, making it similar to the "Average" model.
All five alternatives to the "average" model achieved lower error on some
of the sites in some years, but each one had a higher mean absolute error
and higher mean deviance.

@@

When the differences among observers were not explicitly accounted for, nearly
all metrics of fit worsened (Figure 7). The increased error resulted from a
small number of forecasts where observer turnover caused a large shift in the
reported richness values. The Naive model was less sensitive to these shifts,
because it largely ignored the richness values reported by observers that had
retired by the end of the training period (Figure 1). The Average model, which
gave equal weight to observations from the whole training period, showed a
larger decline in performance -- especially in terms of coverage. The mistnet
JSDM was notable here because the coverage of its prediction intervals retained
very good coverage in the presence of this observer-based noise, which we
attribute to its ability to model this variation with its latent variables.

@@

This research was supported by the Gordon and Betty Moore Foundation's
Data-Driven Discovery Initiative through Grant GBMF4563 to E.P. White. We thank
the developers and providers of the data and software that made this research
possible including: the PRISM Climate Group at Oregon State University, the
staff at USGS and volunteer citizen scientists associated with the North
American Breeding Bird Survey, NASA, the World Climate Research Programme's
Working Group on Coupled Modelling and its working groups, the
U.S. Department of Energy's Program for Climate Model Diagnosis and
Intercomparison, and the Global Organization for Earth System Science Portals.

>>>>

**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
function in @forecast1] to represent a array of different time-series modeling
approaches. These models can include an autoregressive component (as in the
"naive" model, but with the possibility of longer-term dependencies in the
underlying process), a moving average component (where the noise can have serial
autocorrelation) and an integration/differencing component (so that the analysis
could be performed on sequential differences of the raw data, accommodating more
complex patterns including trends). The `auto.arima` function chooses whether to
include each of these components (and how many terms to include for each one)
using AICc [@forecast1]. Since there is no seasonal component to the BBS
time-series, we did not include a season component in these models. Otherwise we
used the default settings for this function [@forecast1].

[[Note: we sometimes refer to the baselines as "time series models", but (1)
we're also using that name for just auto.arima and (2) the "average" model
doesn't use any time series.  Other possible names: site-level models (as
above); richness-only models; environment-free models...]]

### Models: environmental models

In contrast to the site-level time-series models, most attempts to predict
species richness focus on using correlative models based on environmental
variables. We tested three common variants of this approach: direct modeling of
species richness; stacking individual species distribution models; and joint
species distribution models (JSDMs). Following the standard approach, site-level
random effects were not included in these models as predictors, meaning that
this approach implicitly assumes that two sites with identical Bioclim,
elevation, and NDVI values should have identical richness distributions. As
above, we included observer effects and the associated uncertainty by running
these models 500 times (once per MCMC sample).

**"Macroecological" model: richness GBM.** We used a boosted regression tree
model using the `gbm` package [@ridgeway2017] to directly model species richness
as a function of environmental variables. Boosted regression trees are a form of
tree-based modeling that work by fitting thousands of small tree-structured
models sequentially, with each tree optimized to reduce the error of its
predecessors. They are flexible models that are considered well suited for
prediction [@elith2008]. This model was optimized using a Gaussian likelihood,
with a maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000
trees. The number of trees used for prediction was selected using the "out of
bag" estimator; this number averaged `r numbers$n_gbm[1]` for the non-observer
data and `r numbers$n_gbm[2]` for the observer-corrected data. [[add comma separators for thousands place]]

**Species Distribution Model: stacked random forests.** Species distribution
models (SDMs) predict individual species' occurrence probabilities using
environmental variables. Species-level models are used to predict richness by
summing the predicted probability of occupancy across all species at a site.
This avoids known problems with the use of thresholds for determining whether or
not a species will be present at a site [@calabrese2014; @pellissier2013].
Following @calabrese2014, we calculated the uncertainty in our richness estimate
by treating richness as a sum over independent Bernoulli random variables:
$\sigma^2_{richness} = \sum_i{p_i (1 - p_i)}$, where $i$ indexes species. By
itself, this approach is known to underestimate the true community-level
uncertainty because it ignores the uncertainty in the species-level probabilites
[@calabrese2014]. To mitigate this problem, we used an ensemble of 500 estimates
for each of the species-level probabilities instead of just one, propagating the
uncertainty forward. We obtained these estimates using random forests, a common
approach in the species distribution modeling literature. Random forests are
constructed by fitting hundreds of independent regression trees to
randomly-perturbed versions of the data [@cutler2007; @caruana2008]. When
correcting for observer effects, each of the 500 trees in our species-level
random forests used a different Monte Carlo estimate of the observer effects as
a predictor variable.

**Joint Species Distribution Model: mistnet.** Joint species distribution models
(JSDMs) are a new approach that makes predictions about the full composition of
a community instead of modeling each species independently as above. JSDMs
remove the assumed independence among species and explicitly account for the
possibility that a site will be much more (or less) suitable for birds in
general (or particular groups of birds) than one would expect based on the
available environmental measurements alone. As a result, JSDMs do a better job
of representing our uncertainty about richness, while stacked SDMs underestimate
it [@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
[@harris2015] because it is the only JSDM that describes species' environmental
associations with nonlinear functions.

@@

We defined model performance for all models in terms of continuous Gaussian
errors, instead of using discrete count distributions. Variance in species
richness within sites was lower than predicted by several common count models,
such as the Poisson or binomial (i.e. richness was underdispersed for individual
sites), so these count models would have had difficulty fitting the data [cf.
@calabrese2014]. The use of a continuous distribution is adequate here, since
richness had a relatively large mean (`r numbers$richness_summary$mean`) and all
models already produce continuous richness estimates (due to summing SDM or JSDM
probabilities). When a model was run multiple times for the purpose of
correcting for observer effects, we used the mean of those runs' point estimates
as our final point estimate and we calculated the uncertainty using the law of
total variance (i.e. the average of the model runs' variance, plus the variance
in the point estimates).

We evaluated each model's forecasts using the data for each year between `r
timeframe$last_train_year` and `r timeframe$end_yr`. We used three metrics for
evaluating performance: 1) root-mean-square error (RMSE) to determine how far,
on average, the models' predictions were from the observed value; 2) the 95%
prediction interval coverage to determine how well the models predicted the
range of possible outcomes; and 3) deviance (i.e. negative 2 times the Gaussian
log-likelihood) as an integrative measure of fit incorporating good point
estimates, precision, and coverage.  In addition to evaluating forecast
performance in general, we evaluated how performance changed as the time horizon
of forecasting increased by plotting performance metrics against year. Finally,
we decomposed each model's squared error into two components: one described the
squared error associated with site-level means, while the other described the
squared error associated with annual fluctuations in richness within a site.
This decomposition tells us the extent to which each model's error depends on
consistent differences among sites versus changes in site-level richness from
year to year.

All analyses were conducted using R [@R]. Primary R packages used in the
analysis included dplyr [@dplyr], tidyr [@tidyr], gimms [@gimms], sp [@sp1;
@sp2], raster [@raster], prism [@prism], rdataretriever [@rdataretriever],
forecast [@forecast1; @forecast2], git2r [@git2r], ggplot [@ggplot2], mistnet
[@harris2015], viridis [@viridis], rstan [@rstan], yaml [@yaml], purrr [@purrr],
gbm [@gbm], randomForest [@randomForest]. Code to fully reproduce this analysis
is available on GitHub (https://github.com/weecology/bbs-forecasting) and
archived on Zenodo (https://doi.org/10.5281/zenodo.839581).

@@

When comparing forecasts for richness across sites all methods performed well
(Figure 3; all $R^2 > 0.5$). However SDMs (both stacked and joint) and the
macroecological model all failed to successfully forecast the highest-richness
sites, resulting in a notable clustering of predicted values near ~60 species
and the poorest model performance ($R^2$=`r numbers$env_R2s`, versus $R^2$=`r
numbers$ts_R2s` for the within-site methods).

![Performance of the six models we evaluated when predicting one year into the
future (2004) or ten years into the future (2013).  In general, the time series
models (**A.**) outperformed the environmental models. (**B.**) The accuracy of
the predictions generally declined as the forecast horizon was
extended.](../figures/scatter.png)

While the models generally did well in absolute terms (Figure 3), none
consistently outperformed the "average" baseline (Figure 4). The auto-ARIMA was
generally the best-performing non-baseline model, but in many cases (67% of the
time), the auto.arima procedure automatically selected a model with only an
intercept term (i.e. no autoregressive terms, no drift, and no moving average
terms), making it similar to the "average" model. All five alternatives to the
"average" model achieved lower error on some of the sites in some years, but
each one had a higher mean absolute error and higher mean deviance (Figure 4).

@@

When the differences among observers were not explicitly accounted for, nearly
all metrics of fit worsened (Figure 7). The increased error resulted from a
small number of forecasts where observer turnover caused a large shift in the
reported richness values. The naive baseline was less sensitive to these shifts,
because it largely ignored the richness values reported by observers that had
retired by the end of the training period (Figure 1). The average model, which
gave equal weight to observations from the whole training period, showed a
larger decline in performance -- especially in terms of coverage. The
performance of the mistnet JSDM was notable here, because its prediction
intervals retained very good coverage in the presence of uncorrected
observer-based noise, which we attribute to the JSDM's ability to model this
variation with its latent variables.

@@

This research was supported by the Gordon and Betty Moore Foundation's
Data-Driven Discovery Initiative through Grant GBMF4563 to E.P. White. We thank
the developers and providers of the data and software that made this research
possible including: the PRISM Climate Group at Oregon State University, the
staff at USGS and volunteer citizen scientists associated with the North
American Breeding Bird Survey, NASA, the World Climate Research Programme's
Working Group on Coupled Modelling and its working groups, the U.S. Department
of Energy's Program for Climate Model Diagnosis and Intercomparison, and the
Global Organization for Earth System Science Portals. A. C. Perry provided 
valuable comments that improved the clarity of this manuscript.
